---
title: "Hands-on 1"
author: "Martí Sanchis Llovera (marti.sanchos01@estudiant.upf.edu)"
date: 'Last update: `r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    toc: true
    fig_caption: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Load DATA
```{r hd_load}
set.seed(123)
library(pacman)
p_load(zCompositions, here, magrittr, tidyverse, GGally, mice, rcompanion, corrplot, summarytools, gridExtra, patchwork)

dt <- read_delim(file = here("data", "heart_disease_dataset.csv"), delim = " ") %>%
  mutate(patient_id=NULL)

```

## Perform an EDA on the dataset

After loading the dataset with read_delim(), we can see that the dimensions of this tibble are 303 observations with 14 attributes, which are named as specified in the description file. However, there are two variables that don't match the variable type specified in the descriptions: as we see in the header, variables ca and thal are detected as characters but should be categorical variables, and also the rest of categorical variables are detected as floats. Using unique() in these two first commented variables, we see that the reason they are detected as characters is because missing data are coded as a question mark "?". There are no duplicated rows in this dataset.

To correctly parse the variables we use the mutate() function, setting all categorical variables as so (also labeling each factor as described in the description file), and also changing age variable to integer. The categorical variable heart disease diagnosis shows 4 different levels, while in the description the diagnosis can only be positive or negative, so we have changed the levels to positive and negative only. Moreover we rename variables so they have more meaningful names. 

```{r hd_tibble}
dim(dt)
head(dt)
unique(dt$ca)
unique(dt$thal)
# unique(dt$num)
# summary(dt$sex %>% as.factor())
dt[duplicated(dt),]

### Numerical (Discrete/continuous)
# age (numeric discrete): Age in years
# trestbps (numeric): Resting blood pressure (mm Hg)
# chol (numeric): Serum cholesterol (mg/dl)
# thalach (numeric continuous): Maximum heart rate achieved
# oldpeak (numeric continuous): ST depression induced by exercise relative to rest

### Categorical
# sex (factor): 1 = male, 0 = female
# cp: (factor) Chest pain type (1: typical angina, 2: atypical angina, 3: non-anginal pain, 4: asymptomatic)
# fbs (factor): Fasting blood sugar > 120 mg/dl (1 = true, 0 = false)
# restecg (factor): Resting electrocardiographic results (0: normal, 1: ST-T wave abnormality, 2: left ventricular hypertrophy)
# exang (factor): Exercise-induced angina (1 = yes, 0 = no)
# slope (factor): Slope of the peak exercise ST segment (1: upsloping, 2: flat, 3: downsloping)
# ca (factor): Number of major vessels colored by fluoroscopy
# thal (factor): Thalassemia (3: normal, 6: fixed defect, 7: reversible defect)
# num (factor): Diagnosis of heart disease (1: > 50% diameter narrowing, 0: < 50% narrowing)

heart_disease <- dt %>%
  mutate(
    sex = factor(sex, levels = c("0", "1"), labels = c("female", "male")),
    cp = factor(cp, levels = c("1", "2", "3", "4"), labels = c("typical","atypical","non-anginal","asymptomatic")),
    trestbps = as.double(trestbps),
    chol = as.double(chol),
    fbs = factor(fbs, levels = c("0","1"),labels = c("<120mg/dl", ">120mg/dl")),
    restecg = factor(restecg, levels = c("0","1","2"), labels = c("normal", "abnormal", "hypertrophy")),
    thalach = as.double(thalach),
    exang = factor(exang, levels = c("0","1"),labels = c("not", "yes")),
    oldpeak = as.double(oldpeak),
    slope = factor(slope, levels = c("1","2","3"), labels = c("upsloping", "flat", "downsloping")),
    ca = factor(na_if(ca, "?"), levels = c("0","1","2","3")),
    thal = factor(na_if(thal, "?"), levels = c("3","6","7"), labels = c("normal", "fixed", "reversible")),
    num = factor(ifelse(num == "0", "negative", "positive"))
  ) %>% 
    rename(
      Gender=sex,
      Age=age,
      ChestPain = cp,
      RestingBloodPressure = trestbps,
      Cholesterol = chol,
      FastingBloodSugar = fbs,
      RestingECG = restecg,
      MaxHeartRate = thalach,
      ExerciseInducedAngina = exang,
      STDepression = oldpeak,
      Slope=slope,
      MajorVessels = ca,
      Thalassemia = thal,
      Diagnosis = num
  )
head(heart_disease)
summarytools::dfSummary(heart_disease) %>% 
  print(method="render")
```

We use the function zPatterns from the package zCompositions to plot the different patterns of missingness in this dataset. We can observe that very few observations have missing data, only in the two variables mentioned above: vessels_count and thalassemia. Note also that the missingness in the two variables are independent from each other, so most probably these are missing completely at random. Taking into account that only a 2% of observations contain missing data, we could just simply remove these records from the dataset.

We apply the function mice from package "mice" to input the missing data in the two categorical variables. For that, we select the configuration "polyreg" used for imputing missing values for unordered categorical variables with more than 2 levels (non-binary). This method imputes missing data with a Bayesian polytomous regression model. The number of multiple imputations is set to 5 by default. This is the number of datasets generated by the algorithm, imputing different values in each accounting for statistical uncertainty. Later, the algorithm combines these imputations. This algorithm is assigning:
-For MajorVessels: 2 missing values to level 0 and 2 missing values to level 1.
-Thalassemia: 1 missing value to level "normal", and 1 missing value to level "reversible".

```{r hd_missingData}
heart_disease %>% 
  dplyr::select(c(Age,MajorVessels,Thalassemia,Diagnosis)) %>% 
  zPatterns(label = NA, bar.labels = TRUE)
imputed_data <- mice(heart_disease, method = "polyreg",printFlag = FALSE)
heart_disease_imputed <- complete(imputed_data)

summary(heart_disease %>% dplyr::select(MajorVessels,Thalassemia))
summary(heart_disease_imputed %>% dplyr::select(MajorVessels,Thalassemia))
```

Afterwards, we create two different tibbles containing only the numerical variables or the categorical variables. We use the pivot_longer function to reorganize the tibble in a way that the information of attribute is no longer a set of different columns but a single column. The levels of this column are the variable names. The second column contains the corresponding values for all levels. This allows us to make univariate plots for all variables within a single ggplo2 call:

First we plot the histograms from the numerical variables with the density line. The distribution of variable st_depression is heavily zero-inflated and also probably present some outliers in its right tail, deviating the most from a normal distribution. Other variables also present skewed distributions and probably some outliers. Age is the only variable close to a normal distribution.

```{r hd_histograms}
numerical_long <- heart_disease_imputed %>% 
  pivot_longer(cols = where(is.numeric), names_to = "Variables", values_to = "Value") %>% 
  dplyr::select(Variables, Value)

numerical_long %>% ggplot(aes(x = Value)) + 
  geom_histogram(aes(y = after_stat(density)), bins=15, fill = "lightseagreen", color = "black") + 
  geom_density(bw=15) +
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "Histograms for Numeric Variables",
       x = "Value",
       y = "Frequency")
```

Also we use the qq-plots to visualize if they are normally distributed. Variables age, cholesterol, max_hr and rest_bsp align quite well with most of the QQplots, but the tails don't fit really well. Both cholesterol and rest_bps appear to be slightly positively skewed with a heavier right tail and may present some outliers. On the other hand, max_hr shows a slight negative skewness with a heavier left tail. 

```{r hd_qqPlots}
numerical_long %>% ggplot(aes(sample = Value)) +
  geom_qq() +
  geom_qq_line(color = "red") +
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "QQ Plots for Numeric Variables",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")
```

When checking the normality of the numerical variables with the shapiro.wilk test, the assumption of normality is rejected for all of them. We try to use the log values and still normality is rejected for most variables, except from cholesterol.

```{r hd_normalityTest}
numerical_long %>% 
  group_by(Variables) %>%
  summarise(
    p_value = shapiro.test(Value)$p.value)

numerical_long %>% 
  group_by(Variables) %>%
  summarise(
    p_value = shapiro.test(log(Value))$p.value)
```

In the univariate boxplots we can see that some variables like Cholesterol RestingBloodPressure and STDepression have in fact some outliers. In order to adress them, the mahalanobis_distance function from package rstatix is used. This is a multivariate algorithm that uses the mahalanobis distance to flag outliers, also taking into account the covariance matrix of the set of variables. In this way, some of the univariate outliers are not flagged as so. 

```{r hd_univariateBoxplots}
numerical_long %>% ggplot(aes(y = Value)) + 
  geom_boxplot() +
  geom_jitter(aes(x = 0), size=0.8) +
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "Histograms for Numeric Variables",
       x = "Value",
       y = "Frequency")
```

Only observations 153, 127 and 92 are identified as outliers. Looking at the boxplots, is very clear that each of these observations stands out as an outlier (above upper limit) because of its value for a single variable: 153 because of cholesterol value, 127 because of RestingBloodPressure value and 92 because of STDepression value. Consequently, we decide to change a single value in the corresponding variable for each outlier observation using the upper whisker limit of the boxplot, computed as Q3 + 1.5*IQR. After that, the mahalanobis_distance function doesn't detect more outliers.

```{r hd_detectOutliers}
heart_disease_imputed %<>% add_column(rstatix::mahalanobis_distance(heart_disease_imputed) %>%
                                        dplyr::select(is.outlier)) %>% 
  mutate(id=row_number())
  
numerical_long <- heart_disease_imputed %>% 
  pivot_longer(cols = where(is.numeric)& !c("id"), 
               names_to = "Variables", 
               values_to = "Value") %>% 
  dplyr::select(Variables, Value, is.outlier, id)

numerical_long %>% ggplot(aes(y = Value)) + 
  geom_boxplot() +
  geom_jitter(aes(x = 0, colour = is.outlier), size=0.8) +
  geom_text(
    data = numerical_long %>% filter(is.outlier == TRUE),
    aes(x = 0, label = id), 
    hjust = -0.2, 
    size = 3.5, 
    color = "red" 
  ) +
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "Histograms for Numeric Variables",
       x = "Value",
       y = "Frequency") +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red"))

variables_outliers <- list(
  "Cholesterol" = 153,
  "RestingBloodPressure" = 127,
  "STDepression"  = 92)

for (var in names(variables_outliers)) {
  Q1 <- unname(quantile(heart_disease_imputed[[var]], 0.25))
  Q3 <- unname(quantile(heart_disease_imputed[[var]], 0.75))
  IQR <- Q3 - Q1
  upper_limit <- Q3 + 1.5 * IQR
  lower_limit <- Q1 - 1.5 * IQR
  heart_disease_imputed[variables_outliers[[var]],var] <- upper_limit
}

rstatix::mahalanobis_distance(heart_disease_imputed) %>%
  dplyr::select(is.outlier) %>% 
  any()
```

Ending this univariate part, we use barplots for the categorical variables. We can see that many variables' levels are unbalanced. Resting ECG, Slope and Thalassemia have two balanced levels but a third level with very few counts. Gender, Exercise Induced Angina are a little more balanced and the diagnosis variable is well balanced, with almost equal negative and positive diagnosis.

```{r hd_univariateCategorical, echo=FALSE}
plot1 <- ggplot(heart_disease_imputed, aes(x = Gender, fill = Gender)) +
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Gender", x = NULL, y = "")
plot2 <- ggplot(heart_disease_imputed, aes(x = ChestPain, fill = ChestPain)) + 
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1)) +
  labs(title = "Chest Pain", x=NULL, y=NULL)
plot3 <- ggplot(heart_disease_imputed, aes(x = FastingBloodSugar, fill = FastingBloodSugar)) + 
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Fasting Blood Sugar", x=NULL, y=NULL)
plot4 <- ggplot(heart_disease_imputed, aes(x =RestingECG, fill = RestingECG)) +
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Resting ECG",x=NULL,y = "Count")
plot5 <- ggplot(heart_disease_imputed, aes(x = ExerciseInducedAngina, fill = ExerciseInducedAngina)) +
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(title = "Exercice Induced Angina", x = NULL, y=NULL)
plot6 <- ggplot(heart_disease_imputed, aes(x = Slope, fill = Slope)) + 
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") +  
  labs(title = "Slope", x=NULL, y=NULL)
plot7 <- ggplot(heart_disease_imputed, aes(x = MajorVessels, fill = MajorVessels)) + 
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(title = "Major Vessels", x = NULL, y = "")
plot8 <- ggplot(heart_disease_imputed, aes(x = Thalassemia, fill = Thalassemia)) + 
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") +  
  labs(title = "Thalassemia", x = NULL, y=NULL)
plot9 <- ggplot(heart_disease_imputed, aes(x = Diagnosis, fill = Diagnosis)) +  
  geom_bar(color = "black", size = 0.3) + 
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(title = "Diagnosis", x = NULL,y=NULL)

grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot8, plot9, ncol = 3)

```

The scatter plots and correlation coefficients indicate weak relationships between most numerical variables, but a few notable trends emerge:
- Max Heart Rate decreases with Age: This aligns with the physiological decline in cardiovascular efficiency, where the heart's ability to reach high rates diminishes with age.
- ST Depression decreases with Max Heart Rate: Individuals with higher max heart rates may have better cardiovascular health, while those with significant ST depression often have underlying cardiovascular issues.
- Resting Blood Pressure increases with Age: Aging leads to reduced arterial elasticity and increased vascular resistance, contributing to higher blood pressure over time.

```{r hd_ggpairs}
ggpairs(heart_disease_imputed,
        columns = c("Age", "RestingBloodPressure", "Cholesterol", "MaxHeartRate", "STDepression", "Diagnosis"),
        aes(color = Diagnosis),
        progress = FALSE) +  
  theme_minimal() +
  labs(title = "Pair Plot of Numerical Variables by Heart Disease Diagnosis")

```

## Create visualizations in order to show which variables seem to be more associated with heart disease

In order to identify the variables most associated with heart disease, we first compute the correlation matrix between the numerical variables and diagnosis. As diagnosis is a binary categorical variable, we can convert it to numerical and compute the correlation. This would not make sense with a nominal variable with more than two levels, as they wouldn't have any meaningful order. In this case, 0 is negative diagnosis, and 1 is positive diagnosis. Therefore, variables with positive correlation are associated to having the disease, whereas variables with negative correlation are associated with no disease.

As shown below, there are strong correlations among numerical variables and Diagnosis. The most noticeable are Max Heart Rate and ST Depression: 
- MaxHeartRate is negatively correlated with having the disease, which makes sense as a higher heart rate (for instance due to regular exercice) is related to a healthier cardiovascular system
- STDepression is related to cardiovascular issues, and in this dataset is positively correlated with having the disease
As shown in the previous ggpairs plot and in the following boxplots, the most noticeable difference in numerical variables are found in these two variables, which show slightly different distribution dispersion and location when plotted by Diagnosis. Also age is a slightly positively correlated with Diagnosis, suggesting that patients with positive diagnosis tend to be a older.  

```{r hd_numericalCorrelation}
cormat <- cor(heart_disease_imputed %>% 
                select(where(is.numeric), -id, Diagnosis) %>% 
                mutate(Diagnosis = as.numeric(Diagnosis)))
corrplot(cormat, type = "upper", tl.col = "black", tl.srt = 45, addCoef.col = "black", number.cex = 0.7)

boxplot1 <- ggplot(heart_disease_imputed, aes(x=Diagnosis, y=STDepression))+
  geom_violin(alpha = 0.3, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.6) +  
  theme_minimal() +
  labs(title = "ST Depression by diagnosis",
       x = "Diagnosis",
       y = "ST Depression") 

boxplot2 <- ggplot(heart_disease_imputed, aes(x=Diagnosis, y=MaxHeartRate))+
  geom_violin(alpha = 0.3, trim = FALSE) +
  geom_boxplot(width = 0.2, alpha = 0.6) +  
  theme_minimal() +
  labs(title = "Max Heart Rate by diagnosis",
       x = "Diagnosis",
       y = "Max Heart Rate") 
grid.arrange(boxplot1, boxplot2, ncol = 2)

```

In order to measure the association between categorical variables and Diagnosis we use Cramér's V as it is a measure the strength of association between two nominal variables based on the Chi-squared test, allowing a similar view to correlations with numerical variables. For the predictor variables, only Chest Pain and Exercice Induced Angina are associated. Many of the variables seem to be associated with the target variable, mainly Chest Pain, Thalassemia, Major Vessels, Exercice Induced Angina, Slope and Gender. 

```{r hd_categoricalCorrelation}
categorical_vars <- names(heart_disease_imputed %>% dplyr::select(where(is.factor), -id))
cramermat <- matrix(NA, ncol = length(categorical_vars), nrow = length(categorical_vars),
                    dimnames = list(categorical_vars, categorical_vars))
for (i in 1:length(categorical_vars)) {
  for (j in 1:length(categorical_vars)) {
    if (i == j) {
      cramermat[i, j] <- 1
    } else {
      cramermat[i, j] <- cramerV(heart_disease_imputed[[categorical_vars[i]]], 
                                      heart_disease_imputed[[categorical_vars[j]]], 
                                      bias.correct = TRUE)
    }}}
corrplot(cramermat, type = "upper",
         tl.col = "black", tl.srt = 45, 
         addCoef.col = "black",   
         number.cex = 0.7)   
```

Finally, we present the bar plots for the aforementioned categorical variables in relation to Diagnosis:
- Gender: Male patients tend to have a higher proportion of positive diagnoses compared to females, which aligns with the well-documented higher prevalence of heart disease in men.
- Chest Pain Type: Surprisingly, positive diagnoses are most associated with asymptomatic (no chest pain) cases. This suggests that many patients with significant heart disease may not experience typical chest pain.
- Exercise-Induced Angina: Patients who experience angina during exercise show a higher likelihood of a positive diagnosis.
- Slope of the ST Segment: Those with flat or downsloping ST segments are more likely to have heart disease, as these patterns often indicate ischemia cardiovascular issues.
- Major Vessels (colored by fluoroscopy): Patients with one or more colored major vessels tend to have more positive diagnoses.
- Thalassemia: The presence of reversible Thalassemia is strongly associated with positive diagnoses, possibly due to its link with abnormal blood oxygenation and increased cardiac stress.

```{r hd_plots, fig.width=12, fig.height=14}
plot_chisq_fisher_barplot <- function(data, var1, var2) {
  contingency_table <- table(data[[var1]], data[[var2]])
  
  chi_square_result <- chisq.test(contingency_table)
  chi_p_value <- chi_square_result$p.value
  fisher_result <- fisher.test(contingency_table)
  fisher_p_value <- fisher_result$p.value
  
  p <- ggplot(data, aes_string(x = var1, fill = var2)) +
    geom_bar(position = "dodge") +
    labs(title = paste("Barplot of", var1, "by", var2),
         x = var1,
         y = "Frequency") +
    scale_fill_manual(values = c("#1f77b4", "#ff7f0e","darkred","lightgreen")) +  
    theme_minimal() +
    theme(plot.title = element_text(size = 16))  # Ajusta el tamaño del título
  p + annotate("text", x = 1.5, y = max(table(data[[var2]])) * 1.05, 
               label = paste("Chi-sq p-value:", format(chi_p_value, scientific = TRUE, digits = 4)), 
               size = 5, hjust = 0.5, color = "black") +
    annotate("text", x = 1.5, y = max(table(data[[var2]])) * 0.95, 
             label = paste("Fisher p-value:", format(fisher_p_value, scientific = TRUE, digits = 4)), 
             size = 5, hjust = 0.5, color = "black")
}

barplot1 <- plot_chisq_fisher_barplot(heart_disease_imputed, "Gender", "Diagnosis")
barplot2 <- plot_chisq_fisher_barplot(heart_disease_imputed, "ChestPain", "Diagnosis")
barplot3 <- plot_chisq_fisher_barplot(heart_disease_imputed, "ExerciseInducedAngina", "Diagnosis")
barplot4 <- plot_chisq_fisher_barplot(heart_disease_imputed, "Slope", "Diagnosis")
barplot5 <- plot_chisq_fisher_barplot(heart_disease_imputed, "MajorVessels", "Diagnosis")
barplot6 <- plot_chisq_fisher_barplot(heart_disease_imputed, "Thalassemia", "Diagnosis")

(barplot1 + barplot2) / (barplot3 + barplot4) / (barplot5 + barplot6)

```


# 2 Difference in mortality rates in hospitalized COVID-19 patients 
Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}

```


## Reproduce Figure 2 from the publication
but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}

```

## Improve figure 2 of the publication
Add a second annotation with information of deathm and a third one with information of gender

```{r}

```


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
