---
title: "Hands-on 1"
author: "Martí Sanchis Llovera (marti.sanchos01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analysis of the Heart Disease Dataset 
Load the data from
[here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). 
The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Load DATA
```{r hd_load}
library(pacman)
p_load(readr, dplyr, DataExplorer, zCompositions, here, magrittr, ggplot2, tidyr, GGally)

dt <- read_delim(file = here("data", "heart_disease_dataset.csv"), delim = " ") %>%
  mutate(patient_id=NULL)

```

## Perform an EDA on the dataset
```{r hd_eda}
dim(dt)
head(dt)
colnames(dt)

summary(dt)
unique(dt$ca)
unique(dt$thal)
unique(dt$num)

### Numerical (Discrete/continuous)
# age (numeric discrete): Age in years
# trestbps (numeric): Resting blood pressure (mm Hg)
# chol (numeric): Serum cholesterol (mg/dl)
# thalach (numeric continuous): Maximum heart rate achieved
# oldpeak (numeric continuous): ST depression induced by exercise relative to rest

### Categorical
# sex (factor): 1 = male, 0 = female
# cp: (factor) Chest pain type (1: typical angina, 2: atypical angina, 3: non-anginal pain, 4: asymptomatic)
# fbs (factor): Fasting blood sugar > 120 mg/dl (1 = true, 0 = false)
# restecg (factor): Resting electrocardiographic results (0: normal, 1: ST-T wave abnormality, 2: left ventricular hypertrophy)
# exang (factor): Exercise-induced angina (1 = yes, 0 = no)
# slope (factor): Slope of the peak exercise ST segment (1: upsloping, 2: flat, 3: downsloping)
# ca (factor): Number of major vessels colored by fluoroscopy
# thal (factor): Thalassemia (3: normal, 6: fixed defect, 7: reversible defect)
# num (factor): Diagnosis of heart disease (1: > 50% diameter narrowing, 0: < 50% narrowing)

heart_disease <- dt %>%
  mutate(
    sex = factor(sex, levels = c("0", "1"), labels = c("female", "male")),
    cp = factor(cp, levels = c("1", "2", "3", "4"), labels = c("typical","atypical","non-anginal","asymptomatic")),
    trestbps = as.double(trestbps),
    chol = as.double(chol),
    fbs = factor(fbs, levels = c("0","1"),labels = c("<120mg/dl", ">120mg/dl")),
    restecg = factor(restecg, levels = c("0","1","2"), labels = c("normal", "abnormal", "hypertrophy")),
    thalach = as.double(thalach),
    exang = factor(exang, levels = c("0","1"),labels = c("not", "yes")),
    oldpeak = as.double(oldpeak),
    slope = factor(slope, levels = c("1","2","3"), labels = c("upsloping", "flat", "downsloping")),
    ca = factor(na_if(ca, "?"), levels = c("0","1","2","3")),
    thal = factor(na_if(thal, "?"), levels = c("3","6","7"), labels = c("normal", "fixed", "reversible")),
    num = factor(ifelse(num == "0", "negative", "positive"))
  ) %>% 
    rename(
      chest_pain = cp,
      rest_bps = trestbps,
      cholesterol = chol,
      fasting_bs = fbs,
      rest_ecg = restecg,
      max_hr = thalach,
      exercice_angina = exang,
      st_depression = oldpeak,
      slope = slope,
      vessels_count = ca,
      thalassemia = thal,
      hd_diagnosis = num
  )
summary(heart_disease)

numerical_long <- heart_disease %>% 
  pivot_longer(cols = where(is.numeric), names_to = "Variables", values_to = "Value")

categorical_long <- heart_disease %>% 
  pivot_longer(cols = where(is.factor), names_to = "Variables", values_to = "Value")

numerical_long %>% ggplot(aes(x = Value)) + 
  geom_histogram(bins=15, fill = "lightseagreen", color = "black", alpha = 0.7) + 
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "Histograms for Numeric Variables",
       x = "Value",
       y = "Frequency")

numerical_long %>% ggplot(aes(sample = Value)) +
  geom_qq() +
  geom_qq_line(color = "red") +
  facet_wrap(~ Variables, scales = "free") + 
  theme_minimal() +
  labs(title = "QQ Plots for Numeric Variables",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles")


categorical_long %>% ggplot(aes(x = Value, fill = "firebrick")) +
  geom_bar() + 
  facet_wrap(~ Variables, scales = "free") +  
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(title = "Barplots for Categorical Variables",
       x = "Level",
       y = "Frequency")

heart_disease %>% dplyr::select(c(age,sex,chest_pain,vessels_count,thalassemia,hd_diagnosis)) %>% 
  zPatterns(label = NA, bar.labels = TRUE)
plot_missing(heart_disease,
             title = "Missing Data",
             ggtheme = theme_light())

p_ <- GGally::print_if_interactive
ggpairs()

heart_disease %>% dplyr::select(where(is.numeric)) %>% ggpairs()
                         
```


## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r hd_plots}

```


# 2 Difference in mortality rates in hospitalized COVID-19 patients 
Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}

```


## Reproduce Figure 2 from the publication
but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}

```

## Improve figure 2 of the publication
Add a second annotation with information of deathm and a third one with information of gender

```{r}

```


# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
